# OR analysis text output {#quick-start-or-text}



```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(RJafroc)
```

## Introduction {#quick-start-or-text-intro}
This vignette illustrates significance testing using the OR method. But, instead of the unwieldy output in \@ref(quick-start-dbm-text), it generates an Excel output file containing the following worksheets:

* `Summary`
* `FOMs`
* `RRRC`
* `FRRC`
* `RRFC`
* `ANOVA`


## Generating the Excel output file {#quick-start-or-text-output}

This illustrates the `UtilOutputReport()` function. The significance testing `method` is "DBM", the default, and the figure of merit `FOM` is  "Wilcoxon". Note `ReportFileExt` = "xlsx"` telling
the function to create an Excel output file. The Excel output is created in a temporary file.

```{r}
ret <- UtilOutputReport(get("dataset03"), FOM = "Wilcoxon", method = "DBM", overWrite = TRUE, ReportFileExt = "xlsx")
```             


## ORH significance testing {#quick-start-or-text-significance-testing}
Simply change `method = "DBM"` (the default) to `method = "OR"`.


```{r}
ret <- UtilOutputReport(get("dataset03"), FOM = "Wilcoxon", method = "OR", overWrite = TRUE, ReportFileExt = "xlsx")
```             


## References {#quick-start-or-text-references}
