# Optimal operating point using ROC methodology {#optim-op-point-roc}

---
output:
  rmarkdown::pdf_document:
    fig_caption: yes        
    includes:  
      in_header: R/learn/my_header.tex
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval.after = "fig.cap"
)

library(RJafroc)
library(ggplot2)
library(kableExtra)
library(gridExtra)
library(abind)
library(dplyr)
```

## Introduction {#optim-op-point-roc-intro}
Similar to previous chapter except it uses ROC AUC figure of merit.

```{r do_one_mu_roc, echo=FALSE}
do_one_mu_roc <- function(mu, lambda, nu, zeta1Arr, lesDistr, relWeights) {

  fomArray <- array(dim = length(zeta1Arr))
  for (i1 in 1:length(zeta1Arr)) {
    # fomArray[i1] <- UtilAnalyticalAucsRSM(mu, lambda, nu, zeta1 = zeta1Arr[i1], lesDistr, relWeights)$aucwAFROC
    fomArray[i1] <- UtilAnalyticalAucsRSM(mu, lambda, nu, zeta1 = zeta1Arr[i1], lesDistr, relWeights)$aucROC
  }
  # generate plot of AUC vs. zeta1
  myData <- data.frame(zeta1 = zeta1Arr, AUC = fomArray)
  p <- ggplot2::ggplot(myData, aes(x = zeta1, y = AUC)) + geom_line()
  zetaMax <- zeta1Arr[which.max(fomArray)]
  return(list(
    p = p,
    zetaMax = zetaMax,
    maxFom = fomArray[which.max(fomArray)]
  ))
}
```

## Methods {#optim-op-point-roc-methods}

Similar to previous chapter except it uses ROC AUC figure of merit.


### $\zeta_1$ optimization for $\lambda = 10$


Shown next is the variation of ROC AUC vs. $\zeta_1$ for $\lambda = 10$ and the four values of the $\mu$ parameter.


```{r max-fom-arr-10-roc, cache = FALSE, echo=FALSE}
# determine plotArr[[1,]], zetaMaxArr[1,] and maxFomArr[1,]
lambda <- 10
nu <- 1
mu_arr <- c(1, 1.5, 2, 2.5)
maxFomArr <- array(dim = c(2,length(mu_arr)))
zetaMaxArr <- array(dim = c(2, length(mu_arr)))
plotArr <- array(list(), dim = c(2, length(mu_arr)))
lesDistr <- c(0.5, 0.5)
relWeights <- c(0.5, 0.5)
for (i in 1:length(mu_arr)) {
  # if (i == 1) zeta1Arr <- seq(1.5,3.5,0.05) else zeta1Arr <- seq(0.5,2.5,0.1)
  if (i == 1) zeta1Arr <- seq(-3,-1,0.05) else zeta1Arr <- seq(-3,-1,0.1)
  x <- do_one_mu_roc (mu_arr[i], lambda, nu, zeta1Arr, lesDistr, relWeights)
  plotArr[[1,i]] <- x$p + ggtitle(paste0("mu = ", as.character(mu_arr[i]), ", zetaMax = ",  format(x$zetaMax, digits = 3)))
  zetaMaxArr[1,i] <- x$zetaMax
  maxFomArr[1,i] <- x$maxFom
  # plotArr[[2,i]] etc. reserved for lambda = 1 results, done later
}
```


TBA


```{r optim-op-point-roc-AUC-vs-zeta1-10, fig.cap="Variation of AUC vs. $\\zeta_1$ for $\\lambda = 10$; AUC is the wAFROC AUC. panels are labeled by the value of $\\mu$ and zetaMax (i.e., $\\zeta_{\\text{max}}$, the value of $\\zeta_1$ that maximizes AUC).", fig.show='hold', echo=FALSE}
grid.arrange(plotArr[[1,1]],plotArr[[1,2]],plotArr[[1,3]],plotArr[[1,4]], nrow = 2, ncol = 2)
```


Fig. \@ref(fig:optim-op-point-roc-AUC-vs-zeta1-10) corresponding to $\lambda = 10$ has four panels showing the variation of wAFROC AUC with $\zeta_1$. Each panel is labeled by the values of $\mu$ and  $\zeta_{\text{max}}$. For example, the panel labeled `mu = 1.5, zetaMax = 1.9` shows that AUC has a maximum at $\zeta_1 = 1.9$. For $\mu = 1$ the maximum is at the top of a broad plateau, but for higher values the maximum is better defined.


### $\zeta_1$ optimization for $\lambda = 1$

Shown next is the variation of wAFROC AUC vs. $\zeta_1$ for $\lambda = 1$ and the four values of the $\mu$ parameter.

```{r max-fom-arr-01-roc, cache = FALSE, echo=FALSE}
# determine plotArr[[2,]], zetaMaxArr[2,] and maxFomArr[2,]
lambda <- 1

for (i in 1:length(mu_arr)) {
  zeta1Arr <- seq(-1,1,0.05)
  fomArray <- array(dim = length(zeta1Arr))
  x <- do_one_mu_roc (mu_arr[i], lambda, nu, zeta1Arr, lesDistr, relWeights)
  plotArr[[2,i]] <- x$p + ggtitle(paste0("mu = ", as.character(mu_arr[i]), ", zetaMax = ",  format(x$zetaMax, digits = 3)))
  zetaMaxArr[2,i] <- x$zetaMax
  maxFomArr[2,i] <- x$maxFom
}
```



```{r optim-op-point-roc-AUC-vs-zeta1-01, fig.cap="Variation of AUC vs. $\\zeta_1$ for $\\lambda = 1$.", fig.show='hold', echo=FALSE}
grid.arrange(plotArr[[2,1]],plotArr[[2,2]],plotArr[[2,3]],plotArr[[2,4]], nrow = 2, ncol = 2)
```

Fig. \@ref(fig:optim-op-point-roc-AUC-vs-zeta1-01) corresponds to $\lambda = 1$ and employs a similar labeling scheme as Fig. \@ref(fig:optim-op-point-roc-AUC-vs-zeta1-10). For example, the panel labeled `mu = 1, zetaMax = 0.3` shows that AUC has a maximum at $\zeta_1 = 0.3$. 



```{r do_one_roc, echo=FALSE}
do_one_roc <- function(mu, lambda, nu, zeta1, lesDistr, relWeight, llfRange) {
  
  
  froc <- PlotRsmOperatingCharacteristics(
    mu, 
    lambda, 
    nu, 
    zeta1, 
    lesDistr = lesDistr, 
    relWeight = relWeights, 
    OpChType = "FROC",
    llfRange = llfRange,
    legendPosition = "none")$FROCPlot
  
  return(froc)
}
```



```{r detOptOpPt-roc, cache = FALSE, echo=FALSE}
# determine optOpPt here

lambdaArr <- c(10,1)
nu <- 1
#mu_arr <- c(1, 1.5, 2, 2.5) #already defined above
#zetaMaxArr                  #already defined above
plotArr <- array(list(), dim = c(2,length(mu_arr)))

# indices: lambda, nlf/llf, 4-values of mu
OpPtMax <- optOpPt <- array(dim = c(2, 2, 4))

for (j in 1:2) {
  for (i in 1:length(mu_arr)) {
    if ((j ==1) && (i ==1)) llfRange <- c(0, 1) else llfRange <- c(0,1) 
    x <- do_one_roc (mu_arr[i],
                      lambdaArr[j],
                      nu,
                      zeta1 = zetaMaxArr[j,i],
                      lesDistr, 
                      relWeight, 
                      llfRange = NULL)
    # optimal operating point
    optOpPt[j,1,i] <- max(x$data$NLF)
    optOpPt[j,2,i] <- max(x$data$LLF)
    
    # following are end-points assuming zeta1 = -Inf
    physicalValues <- UtilIntrinsic2PhysicalRSM(mu_arr[i], lambdaArr[j], nu)
    # OpPtMax[j,1,i] <- physicalValues$lambdaP*pnorm(-zetaMaxArr[j,i])
    # OpPtMax[j,2,i] <- physicalValues$nuP*pnorm(mu_arr[i]-zetaMaxArr[j,i])
    OpPtMax[j,1,i] <- physicalValues$lambdaP
    OpPtMax[j,2,i] <- physicalValues$nuP
  }
}
```



```{r useOptOpPt-roc, cache = FALSE, echo=FALSE}

# use optOpPt here
# generate plotArr

lambdaArr <- c(10,1)
nu <- 1
# following are coordinates of extended plots
OpPtExtended <- array(dim = c(2, 2, 4))
# optOpPt defined above
#mu_arr <- c(1, 1.5, 2, 2.5) #already defined above
#zetaMaxArr                  #already defined above
plotArr <- array(list(), dim = c(2,length(mu_arr)))

correction <- array(0.5, dim = c(2, 4))

for (j in 1:2) {
  for (i in 1:length(mu_arr)) {
    if ((j ==1) && (i ==1)) llfRange <- c(0, 1) else llfRange <- c(0,1) 
    x <- do_one_roc (
      mu_arr[i],
      lambdaArr[j],
      nu,
      # to show a fuller extent of the plot
      zeta1 = zetaMaxArr[j,i] - correction[j,i],
      lesDistr, 
      relWeight, 
      llfRange)

    plotArr[[j,i]] <- x +
      geom_vline(
        xintercept =
          optOpPt[j,1,i], color = "blue")  +
      ggtitle(
        paste0(
          as.character(lambdaArr[j]),
          "-",
          as.character(mu_arr[i])))

    OpPtExtended[j,1,i] <- max(x$data$NLF)
    OpPtExtended[j,2,i] <- max(x$data$LLF)
  }
}
```



```{r optim-op-point-roc-cad-optim-zeta1, echo=FALSE}
x1 <- as.data.frame(rbind(t(maxFomArr[1,]), t(optOpPt[1,1,]), t(optOpPt[1,2,])))
x2 <- as.data.frame(rbind(t(maxFomArr[2,]), t(optOpPt[2,1,]), t(optOpPt[2,2,])))
x12 <- rbind(x1, x2)
x12 <- cbind(c("AUC10", "NLF10", "LLF10", "AUC01", "NLF01", "LLF01"), x12)
colnames(x12) <- c("measure", paste0("mu", as.character(mu_arr)))
y <- c("AUC10", "AUC01", "NLF10", "LLF10", "NLF01", "LLF01")
x12 %>% slice(match(y, measure)) -> xFinal
```


### Summary of simulations and comments {#optim-op-point-roc-comments-threshold-optimization}


```{r optim-op-point-roc-cad-optim-table-roc, echo=FALSE}
knitr::kable(xFinal, digits = 5, caption = "Summary of optimal threshold values: \"measure\" refers to a performance measure, \"mu1\" refers to $\\mu = 1$, etc. The row labels are as follows: AUC10 is the wAFROC AUC for lambda = 10, AUC01 is the wAFROC AUC for $\\lambda = 1$, NLF10 is NLF for $\\lambda = 10$, ..., LLF01 is LLF for $\\lambda = 1$.", escape = FALSE)
```

TBA

#### Explanations {#optim-op-point-roc-threshold-explanations}

```{r optim-op-point-roc-roc-10-first-two-plots, fig.cap="Extended FROC plots: panel labeled 10-1 is for $\\lambda = 10$ and $\\mu = 1$, and that labeled 10-1.5 is for $\\lambda = 10$ and $\\mu = 1.5$. The blue line indicates the optimal operating point.", fig.show='hold', echo=FALSE, fig.height=3.5,  fig.pos='H'}
grid.arrange(plotArr[[1,1]], plotArr[[1,2]], nrow = 1, ncol = 2)
```             


TBA



```{r optim-op-point-roc-roc-10-next-two-plots, fig.cap="Extended FROC plots: panel labeled 10-2 is for $\\lambda = 10$ and $\\mu = 2$ and that labeled 10-2.5 is for $\\lambda = 10$ and $\\mu = 2.5$. The blue line indicates the optimal operating point.", fig.show='hold', echo=FALSE, fig.height=3.5,  fig.pos='H'}
grid.arrange(plotArr[[1,3]], plotArr[[1,4]], nrow = 1, ncol = 2)
```


TBA


```{r optim-op-point-roc-roc-01-first-two-plots, fig.cap="Extended FROC plots: panel labeled 1-1 is for $\\lambda = 1$ and $\\mu = 1$ and that labeled 10-1.5 is for $\\lambda = 1$ and $\\mu = 1.5$. The blue line indicates the optimal operating point.", fig.show='hold', echo=FALSE, fig.height=3.5,  fig.pos='H'}
grid.arrange(plotArr[[2,1]], plotArr[[2,2]], nrow = 1, ncol = 2)
```             


TBA

```{r optim-op-point-roc-roc-01-next-two-plots, fig.cap="Extended FROC plots: panel labeled 1-2 is for $\\lambda = 1$ and $\\mu = 2$ and that labeled 1-2.5 is for $\\lambda = 1$ and $\\mu = 2.5$. The blue line indicates the optimal operating point.", fig.show='hold', echo=FALSE, fig.height=3.5,  fig.pos='H'}
grid.arrange(plotArr[[2,3]], plotArr[[2,4]], nrow = 1, ncol = 2)
```



* In Fig. \@ref(fig:optim-op-point-roc-roc-01-next-two-plots) panel labeled **1-2**: The vertical blue line is at NLF = `r format(optOpPt[2,1,3], digits = 3)` corresponding to LLF = `r format(optOpPt[2,2,3], digits = 3)`. The end-point of the extended curve is (NLF = `r format(OpPtExtended[2,1,3], digits = 3)`, LLF = `r format(OpPtExtended[2,2,3], digits = 3)`). The highest operating point is at (NLF = `r format(OpPtMax[2,1,3], digits = 3)`, LLF = `r format(OpPtMax[2,2,3], digits = 3)`). 


* In Fig. \@ref(fig:optim-op-point-roc-roc-01-next-two-plots) panel labeled **1-2.5**: The vertical blue line is at NLF = `r format(optOpPt[2,1,4], digits = 3)` corresponding to LLF = `r format(optOpPt[2,2,4], digits = 3)`. The end-point of the extended curve is (NLF = `r format(OpPtExtended[2,1,4], digits = 3)`, LLF = `r format(OpPtExtended[2,2,4], digits = 3)`). The highest operating point is at (NLF = `r format(OpPtMax[2,1,4], digits = 3)`, LLF = `r format(OpPtMax[2,2,4], digits = 3)`). 



## How to use the method {#optim-op-point-roc-how-to-use-method}
TBA  



## Discussion {#optim-op-point-roc-Discussion}
TBA 

## References {#optim-op-point-roc-references}
